---
title: "PCA"
output: html_notebook
---

Let's first create a synthetic data to play with.

```{r}
x <- sin(seq(0, 6.5*pi, length.out = 360))
y <- sin(seq(pi/16, 6.5*pi, length.out = 365))
z <- jitter(y, amount = 0.5) # jitter data the amount of 0.5
t <- z[6:365] # lag 5 units
y <- y[1:360]; z <- z[1:360]
k <- x + cos(seq(0, 128*pi, length.out = 360)) * 0.2
df <- data.frame(x = x, y = y, z = z, t = t, k = k)
pca <- prcomp(df, center = TRUE, scale. = TRUE)
```

```{r, echo=FALSE}
matplot(df, type = "l", bty = "n")
plot(df, pch = ".", cex = 0.1)
```

Basic statistics:

```{r}
round(cor(df), 2) # correlation
round(var(df), 2) # variance & covariance
round(sqrt(diag(var(df))), 2) # std dev
```

### Variances of the principal components

```{r}
eig <- (pca$sdev) ^ 2 # Eigenvalues
print(eig)
variance <- eig * 100/sum(eig) # variance in percentage
print(variance)
cumvar <- cumsum(variance) # Cumulative variance
print(cumvar)
```

### Coordinates of variables on the principal components

The correlation between variables and principal components

```{r}
var_cor_func <- function(var.loadings, comp.sdev) var.loadings*comp.sdev
# Variable correlation/coordinates
loadings <- pca$rotation
sdev <- pca$sdev
var.coord <- var.cor <- t(apply(loadings, 1, var_cor_func, sdev))
head(var.coord[, 1:4])
```

```{r}
print(pca)
s.pca <- summary(pca) # variation explained by each component (as percentage)
print(s.pca)
layout(matrix(c(1,2,3,4), 2, 2, byrow = TRUE), widths = c(2,2), heights = c(2,2))
plot(pca) # Variance explained by each component
# or let's plot proportion of variance as percentage
plot(s.pca$importance[2,]*100, type = "b", bty = "n", ylab = "proportion of variance (%)", xlab = "PC")
screeplot(pca, type = "lines") # another function to plot variation of variance
abline(h = 1,lty = 3, col = "red")
biplot(pca)
```

#### How many values to keep?

* Kaiser Criterion: retain only factors with eigenvalues > 1 (Note: eigenvalues = variances)

### Homemade biplot

```{r}
myplot <- function(r, d1 = 1, d2 = 2) {
  r <- r[, c(d1, d2)]
  mx <- max(range(r[,1]))
  plot(r, type = "n", ylim = c(-1, 1), xlim = c(-mx*1.2, mx*1.2), xlab = paste0("PC", d1), ylab = paste0("PC", d2))
  abline(v = 0, h = 0, lty = 3)
  arrows(0, 0, r[,1], r[,2], len = 0.1, col = "darkred")
  text(1.1 * r, rownames(r), col = "darkred", xpd = TRUE, cex = 1)
}
r <- pca$rotation
layout(matrix(c(1,2,3,4,5,6), 2, 3, byrow = TRUE), widths = c(1,1,1), heights = c(1,1,1))
myplot(r)
myplot(r, 2, 3)
myplot(r, 1, 3)
myplot(r, 3, 4)
myplot(r, 4, 5)
myplot(r, 2, 5)
```

```{r}
library(scatterplot3d)
mx1 <- max(range(r[,1]))
mx2 <- max(range(r[,2]))
mx3 <- max(range(r[,3]))
plt <- scatterplot3d(r[,1], r[,2], r[,3], angle = 130, asp = 0.5,
              xlim = c(-mx1, mx1), ylim = c(-mx2, mx2), zlim = c(-mx3, mx3),
              xlab = "PC1", ylab = "PC2", zlab = "PC3")
plt$points3d(x=c(0,0), y=c(0,0),z=c(-mx3*1.2, mx3), type="l", col="black", lwd = 2)
plt$points3d(x=c(-mx1,mx1), y=c(0,0),z=c(0,0), type="l", col="black", lwd = 2)
plt$points3d(x=c(0,0), y=c(-mx2,mx2),z=c(0,0), type="l", col="black", lwd = 2)
for (i in 1:5)
  plt$points3d(x=c(0,r[i,1]), y=c(0,r[i,2]),z=c(0, r[i,3]), type="l", col="blue", lwd=2)
```

### PC Calculation

```{r}
pc1 <- drop(scale(as.matrix(df), center = pca$center,
             scale = pca$scale) %*% pca$rotation[, 1])
all.equal(pc1, pca$x[,1]) # TRUE
```

### Contributions of the variables to the principal components

```{r}
var.cos2 <- var.coord^2
comp.cos2 <- apply(var.cos2, 2, sum)
contrib <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))
print(var.contrib)
colSums(var.contrib)
```


## RESOURCES:

* http://rcourse.iop.kcl.ac.uk/2014/5fri/3/updated_kh_pca_slides_030614.pdf
* https://cran.r-project.org/web/packages/HSAUR/vignettes/Ch_principal_components_analysis.pdf
* https://stats.stackexchange.com/questions/262611/r-pca-principal-psych-package-vs-prcomp-loadings
* http://www.sthda.com/english/wiki/principal-component-analysis-in-r-prcomp-vs-princomp-r-software-and-data-mining


